---
title: "Case Study — Lung Adenocarcinoma Serum/Plasma Metabolomics: QC & Differential Analysis"
author: "Santiago López Begines"
date: "`r Sys.Date()`"
format:
  html:
    theme: lux
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    code-line-numbers: true
    fig-width: 10
    fig-height: 8
    fig-cap-location: bottom
    highlight-style: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This notebook is **Track A (univariate differential analysis)** of the biomarker case
study (see `docs/case_study_ML_biomarker_design.md`). It performs quality control,
normalisation, and differential analysis of **Adenocarcinoma vs Healthy** using standard
Bioconductor and CRAN packages (limma, ggplot2, pheatmap, factoextra).

Data come from two GC-TOF-MS cohorts of project PR000293:
**ADC1 = ST000368** (training) and **ADC2 = ST000369** (independent validation).
Per the design, the differential is run **within ADC1 only**, **separately per biofluid**
(Plasma, Serum), on the **137 metabolites shared** by both cohorts.

**Normalisation pipeline:** median row-normalisation → Log2 transform → Pareto scaling
(PCA only). **DE:** limma moderated t-test, model `~ class + smoker + gender`, FDR
(Benjamini–Hochberg), threshold `adj.P < 0.05` and `|log2FC| > log2(FC_THRESH)`.

# Setting up environment

```{r env}
source(file.path("..", "config", "params.R"))

set.seed(PARAMS$seed)

# Differential thresholds
FC_THRESH   <- 1.2
PVAL_THRESH <- 0.05

# Libraries
source(file.path(PROJ, "code", "00_packages.R"))

# Normalisation + limma helpers (this analysis)
source(file.path(PROJ, "code", "01_aux_functions.R"))
source(file.path(PROJ, "code", "02_aux_metabolomics.R"))

# Global output state used by save_plot() / save_table() / save_checkpoint()
output_path  <- NB_OUT
image_number <- 1

# MetaboAnalystR is kept as optional; this notebook does not require it
HAS_METABO <- requireNamespace("MetaboAnalystR", quietly = TRUE)
if (HAS_METABO) suppressMessages(library(MetaboAnalystR))
if (!exists("dpi")) dpi <- 150

cat("MetaboAnalystR available:", HAS_METABO, "\n")
cat("Output path:", output_path, "\n")
```

# Loading data

```{r build-matrices}
source(file.path(PROJ, "code", "03_build_matrices.R"))
```

Pre-built analysis matrices are produced by `code/03_build_matrices.R`.
Each file holds samples in rows: metadata columns
(`sample_id`, `cohort`, `biofluid`, `class`, `smoker`, `gender`) followed by the
137 shared metabolites.

```{r load}
read_matrix <- function(name)
  read.csv(file.path(DATA, paste0(name, ".csv")), check.names = FALSE)

mats <- list(
  ADC1_Plasma = read_matrix("ADC1_Plasma"),
  ADC1_Serum  = read_matrix("ADC1_Serum"),
  ADC2_Plasma = read_matrix("ADC2_Plasma"),
  ADC2_Serum  = read_matrix("ADC2_Serum")
)
META_COLS <- c("sample_id", "cohort", "biofluid", "class", "smoker", "gender")
FEATURES  <- setdiff(colnames(mats$ADC1_Plasma), META_COLS)
cat("Shared metabolites (features):", length(FEATURES), "\n")
```

# Experimental design

```{r design}
comp <- do.call(rbind, lapply(mats, function(d)
  as.data.frame(table(cohort = d$cohort, biofluid = d$biofluid, class = d$class))))
comp <- comp[comp$Freq > 0, ]
knitr::kable(comp, caption = "Composition: Adenocarcinoma vs Healthy by cohort and biofluid") |>
  kableExtra::kable_paper(full_width = FALSE, position = "center")
```

The differential (Track A) uses the **ADC1** cohort only, one model per biofluid.

```{r clean-env}
rm(list = intersect(
  c("adc1","adc2","fr","matrices","mdata.all","mdata.siggenes",
    "obj","sample_sheet","shared_tbl"),
  ls()
))
```

# Quality Control

## Missing values

```{r missing-values}
convert_to_factor <- function(df) {
  df |>
    dplyr::mutate(
      cohort   = as.factor(cohort),
      biofluid = as.factor(biofluid),
      class    = as.factor(class),
      smoker   = as.factor(smoker),
      gender   = as.factor(gender)
    )
}
mats <- lapply(mats, convert_to_factor)

cat("Missing values present:\n")
for (nm in names(mats)) cat(" ", nm, ":", anyNA(mats[[nm]]), "\n")

na_summary <- function(df) {
  df |>
    dplyr::summarise(dplyr::across(dplyr::everything(),
                                    \(x) round(mean(is.na(x)) * 100, 2))) |>
    tidyr::pivot_longer(dplyr::everything(),
                        names_to = "variable", values_to = "pct_na") |>
    dplyr::filter(pct_na > 0) |>
    dplyr::arrange(dplyr::desc(pct_na))
}

NA_distribution <- dplyr::bind_rows(lapply(mats, na_summary), .id = "matrix")
NA_distribution |>
  knitr::kable(caption = "Missing values (%) per variable and matrix") |>
  kableExtra::kable_paper(full_width = FALSE, position = "center")

# kNN imputation (k = 10) on numeric feature columns only
matrices_with_na <- unique(NA_distribution$matrix)
mats[matrices_with_na] <- lapply(mats[matrices_with_na], function(df) {
  meta   <- dplyr::select(df, dplyr::where(Negate(is.numeric)))
  feats  <- dplyr::select(df, dplyr::where(is.numeric))
  imputed <- tibble::as_tibble(VIM::kNN(feats, k = 10, imp_var = FALSE))
  dplyr::bind_cols(meta, imputed)
})

NA_distribution_post <- dplyr::bind_rows(lapply(mats, na_summary), .id = "matrix")
if (nrow(NA_distribution_post) == 0) {
  cat("All missing values successfully imputed.\n")
} else {
  NA_distribution_post |>
    knitr::kable(caption = "Remaining missing values after imputation") |>
    kableExtra::kable_paper(full_width = FALSE, position = "center")
}
```

## Feature preparation

Restrict to ADC1. Remove features with constant or near-zero variance (CV < 10 %).

```{r feature-prep}
# ADC1 only
adc1 <- list(
  ADC1_Plasma = mats$ADC1_Plasma,
  ADC1_Serum  = mats$ADC1_Serum
)

# Extract feature sub-matrices to check variance before normalisation
feat_counts <- lapply(names(adc1), function(nm) {
  df        <- adc1[[nm]]
  feat_cols <- intersect(FEATURES, colnames(df))
  raw_mat   <- as.matrix(df[, feat_cols, drop = FALSE])
  storage.mode(raw_mat) <- "double"
  cv  <- apply(raw_mat, 2, function(x) sd(x, na.rm=TRUE) / abs(mean(x, na.rm=TRUE)))
  n_low_var <- sum(cv < 0.10, na.rm = TRUE)
  data.frame(
    matrix          = nm,
    n_samples       = nrow(raw_mat),
    n_features_raw  = ncol(raw_mat),
    n_low_var_removed = n_low_var,
    n_features_kept = ncol(raw_mat) - n_low_var
  )
})

do.call(rbind, feat_counts) |>
  knitr::kable(caption = "Feature count before/after low-variance filter") |>
  kableExtra::kable_paper(full_width = FALSE, position = "center")
```

# Normalisation

## Pipeline

Median row normalisation → Log2 transform → Pareto scaling (PCA only).
The log2 transform is used so that limma's logFC is directly in log2 units
(threshold: `|log2FC| > log2(FC_THRESH)` ≈ `r round(log2(FC_THRESH), 3)`).

```{r normalise, cache=TRUE}
if (check_checkpoint("norm_results")) {
  prep_res <- load_checkpoint("norm_results")
} else {
  prep_res <- lapply(names(adc1), function(nm) {
    feat_cols <- intersect(FEATURES, colnames(adc1[[nm]]))
    prepare_matrix(adc1[[nm]], feat_cols = feat_cols, meta_cols = META_COLS)
  })
  names(prep_res) <- names(adc1)
  save_checkpoint(prep_res, "norm_results")
}

# Dimensions after normalisation
for (nm in names(prep_res)) {
  cat(nm, "— DE matrix:", nrow(prep_res[[nm]]$mat_de),
      "samples ×", ncol(prep_res[[nm]]$mat_de), "features\n")
}
```

## Normalisation QC plots

Density (all features, per sample) and per-sample boxplots before and after log2 transform.

```{r norm-qc-plots, fig.width=14, fig.height=10}
make_norm_qc <- function(raw_mat, log2_mat, meta, label) {
  cls <- as.character(meta$class[match(rownames(raw_mat), meta$sample_id)])

  # Long format helpers
  to_long <- function(mat, stage) {
    df <- as.data.frame(mat)
    df$sample <- rownames(df)
    df$class  <- cls
    df$stage  <- stage
    tidyr::pivot_longer(df, -c(sample, class, stage),
                        names_to = "feature", values_to = "value")
  }

  raw_long  <- to_long(raw_mat,  "Raw (imputed)")
  log2_long <- to_long(log2_mat, "Log2")

  # Random 30 features for boxplots
  set.seed(PARAMS$seed)
  feat_sub  <- sample(colnames(raw_mat), min(30L, ncol(raw_mat)))

  # Density of sample means (before vs after)
  mean_df <- dplyr::bind_rows(
    raw_long  |> dplyr::summarise(mean_val = mean(value, na.rm=TRUE),
                                   .by = c(sample, class, stage)),
    log2_long |> dplyr::summarise(mean_val = mean(value, na.rm=TRUE),
                                   .by = c(sample, class, stage))
  ) |> dplyr::mutate(stage = factor(stage, levels = c("Raw (imputed)", "Log2")))

  p_density <- ggplot2::ggplot(mean_df,
                                ggplot2::aes(x = mean_val, colour = class,
                                             linetype = stage)) +
    ggplot2::geom_density(linewidth = 0.9) +
    ggplot2::scale_colour_manual(values = c(Adenocarcinoma = "firebrick3",
                                             Healthy = "steelblue4")) +
    ggplot2::facet_wrap(~stage, scales = "free_x") +
    ggplot2::labs(title = paste("Sample-mean density —", label),
                  x = "Mean intensity", y = "Density",
                  colour = "Class", linetype = "Stage") +
    ggplot2::theme_bw()

  # Per-sample boxplot after log2 (coloured by class)
  log2_sub  <- log2_long |> dplyr::filter(feature %in% feat_sub)
  p_box <- ggplot2::ggplot(log2_sub,
                            ggplot2::aes(x = sample, y = value, fill = class)) +
    ggplot2::geom_boxplot(outlier.size = 0.3, linewidth = 0.3) +
    ggplot2::scale_fill_manual(values = c(Adenocarcinoma = "firebrick3",
                                           Healthy = "steelblue4")) +
    ggplot2::labs(title = paste("Per-sample distribution (log2, 30 features) —", label),
                  x = NULL, y = "log2 intensity", fill = "Class") +
    ggplot2::theme_bw() +
    ggplot2::theme(axis.text.x = ggplot2::element_blank(),
                   axis.ticks.x = ggplot2::element_blank())

  patchwork::wrap_plots(p_density, p_box, ncol = 1, heights = c(1, 1.5))
}

for (nm in names(prep_res)) {
  p <- make_norm_qc(
    raw_mat  = prep_res[[nm]]$raw,
    log2_mat = prep_res[[nm]]$mat_de,
    meta     = prep_res[[nm]]$meta,
    label    = nm
  )
  print(p)
  save_plot(paste0("norm_qc_", nm), p, width = 12, height = 9, subdir = "QC")
}
```

# Exploratory Data Analysis

## Feature distributions

```{r eda-distributions, fig.width=12, fig.height=8}
# Top 30 variable features per matrix (Pareto-scaled, for visualisation)
for (nm in names(prep_res)) {
  var_rank <- sort(apply(prep_res[[nm]]$mat_pca, 2, var, na.rm=TRUE),
                   decreasing = TRUE)
  top30    <- names(var_rank)[seq_len(min(30L, length(var_rank)))]

  long_df <- as.data.frame(prep_res[[nm]]$mat_pca[, top30]) |>
    tibble::rownames_to_column("sample_id") |>
    dplyr::left_join(
      dplyr::select(prep_res[[nm]]$meta, sample_id, class),
      by = "sample_id"
    ) |>
    tidyr::pivot_longer(-c(sample_id, class),
                        names_to = "feature", values_to = "value") |>
    dplyr::mutate(feature = stringr::str_trunc(feature, 20))

  p <- ggplot2::ggplot(long_df,
                        ggplot2::aes(x = value, fill = class)) +
    ggplot2::geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
    ggplot2::scale_fill_manual(values = c(Adenocarcinoma = "firebrick3",
                                           Healthy = "steelblue4")) +
    ggplot2::facet_wrap(~feature, scales = "free", ncol = 6) +
    ggplot2::labs(title = paste("Top 30 variable features (Pareto-scaled) —", nm),
                  x = "Pareto-scaled log2 intensity", y = "Count", fill = "Class") +
    ggplot2::theme_bw(base_size = 8) +
    ggplot2::theme(legend.position = "bottom")

  print(p)
  save_plot(paste0("eda_distributions_", nm), p,
            width = 14, height = 10, subdir = "EDA")
}
```

## Feature correlation

```{r eda-correlation, fig.width=10, fig.height=9}
for (nm in names(prep_res)) {
  var_rank <- sort(apply(prep_res[[nm]]$mat_de, 2, var, na.rm=TRUE),
                   decreasing = TRUE)
  top30 <- names(var_rank)[seq_len(min(30L, length(var_rank)))]

  p_corr <- DataExplorer::plot_correlation(
    as.data.frame(prep_res[[nm]]$mat_de[, top30]),
    title = paste("Feature correlation (top 30, log2) —", nm)
  )
  print(p_corr)
  save_plot(paste0("eda_correlation_", nm), p_corr,
            width = 10, height = 9, subdir = "EDA")
}
```

# Principal Component Analysis

```{r pca-analysis, fig.width=14, fig.height=12}
pca_results <- list()

for (nm in names(prep_res)) {
  mat_pca <- prep_res[[nm]]$mat_pca
  meta    <- prep_res[[nm]]$meta
  cls     <- meta$class[match(rownames(mat_pca), meta$sample_id)]

  pca_obj <- prcomp(mat_pca, center = FALSE, scale. = FALSE)
  pca_results[[nm]] <- pca_obj

  # Variance explained
  pct_var <- round(100 * pca_obj$sdev^2 / sum(pca_obj$sdev^2), 1)

  # Scree plot
  p_scree <- factoextra::fviz_eig(
    pca_obj, addlabels = TRUE, ncp = min(10L, ncol(mat_pca)),
    barfill = "steelblue4", barcolor = "steelblue4"
  ) +
    ggplot2::labs(title = paste("Scree plot —", nm)) +
    ggplot2::theme_bw()

  # Score plot (PC1 × PC2)
  p_score <- factoextra::fviz_pca_ind(
    pca_obj,
    geom.ind     = "point",
    col.ind      = cls,
    palette      = c(Adenocarcinoma = "firebrick3", Healthy = "steelblue4"),
    addEllipses  = TRUE,
    ellipse.level = 0.95,
    legend.title = "Class",
    title        = paste("PCA Score Plot (PC1 × PC2) —", nm)
  ) +
    ggplot2::labs(
      x = paste0("PC1 (", pct_var[1], "%)"),
      y = paste0("PC2 (", pct_var[2], "%)")
    ) +
    ggplot2::theme_bw()

  # Biplot (top 15 loading vectors)
  p_biplot <- factoextra::fviz_pca_biplot(
    pca_obj,
    geom.ind     = "point",
    col.ind      = cls,
    col.var      = "grey40",
    palette      = c(Adenocarcinoma = "firebrick3", Healthy = "steelblue4"),
    addEllipses  = TRUE,
    ellipse.level = 0.95,
    select.var   = list(contrib = 15),
    repel        = TRUE,
    label        = "var",
    legend.title = "Class",
    title        = paste("PCA Biplot (top 15 loadings) —", nm)
  ) +
    ggplot2::labs(
      x = paste0("PC1 (", pct_var[1], "%)"),
      y = paste0("PC2 (", pct_var[2], "%)")
    ) +
    ggplot2::theme_bw()

  combined <- (p_scree + p_score) / p_biplot +
    patchwork::plot_annotation(title = nm,
                               theme = ggplot2::theme(
                                 plot.title = ggplot2::element_text(size = 14,
                                                                     face = "bold")))
  print(combined)
  save_plot(paste0("pca_", nm), combined, width = 14, height = 12, subdir = "PCA")
}
```

# Covariate Assessment

Balance check and aliasing test for `smoker` and `gender` vs `class`.

```{r covariate-check}
for (nm in names(prep_res)) {
  cat("\n===", nm, "===\n")
  check_covariates(prep_res[[nm]]$meta)
}
```

# Differential Analysis

## limma model: `~ class + smoker + gender`

```{r limma-de, cache=TRUE}
if (check_checkpoint("de_results")) {
  de_results <- load_checkpoint("de_results")
} else {
  de_results <- lapply(names(prep_res), function(nm) {
    run_limma_de(
      mat_de = prep_res[[nm]]$mat_de,
      meta   = prep_res[[nm]]$meta,
      label  = nm
    )
  })
  names(de_results) <- names(prep_res)
  save_checkpoint(de_results, "de_results")
}
```

## Results tables

```{r de-tables}
fc_thresh_log2 <- log2(FC_THRESH)

for (nm in names(de_results)) {
  de <- de_results[[nm]]
  n_sig <- sum(de$significant, na.rm = TRUE)
  n_up  <- sum(de$significant & de$logFC > 0, na.rm = TRUE)
  n_dn  <- sum(de$significant & de$logFC < 0, na.rm = TRUE)

  cat(sprintf("\n%s — %d significant (↑%d, ↓%d) at FDR < %.2f & |log2FC| > %.3f\n",
              nm, n_sig, n_up, n_dn, PVAL_THRESH, fc_thresh_log2))

  de |>
    dplyr::filter(significant) |>
    dplyr::select(metabolite, logFC, t_stat, p_value, adj_p_value) |>
    dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 4))) |>
    knitr::kable(caption = paste("Significant metabolites —", nm)) |>
    kableExtra::kable_paper(full_width = FALSE, position = "center") |>
    print()

  save_table(de, paste0("limma_de_", nm), subdir = "DE/")
}
```

## Summary across matrices

```{r de-summary}
all_de <- dplyr::bind_rows(de_results)

summary_tbl <- all_de |>
  dplyr::group_by(label) |>
  dplyr::summarise(
    n_tested    = dplyr::n(),
    n_sig       = sum(significant, na.rm = TRUE),
    n_up        = sum(significant & logFC > 0, na.rm = TRUE),
    n_down      = sum(significant & logFC < 0, na.rm = TRUE),
    top_up      = metabolite[which.max(ifelse(significant, logFC,  -Inf))],
    top_down    = metabolite[which.min(ifelse(significant, logFC,   Inf))],
    .groups     = "drop"
  )

summary_tbl |>
  knitr::kable(caption = "Differential analysis summary (ADC1 cohort)") |>
  kableExtra::kable_paper(full_width = FALSE, position = "center")
```

# Visualisation

## Volcano plots

```{r volcano-plots, fig.width=12, fig.height=6}
make_volcano <- function(de, label) {
  fc_thresh_log2 <- log2(FC_THRESH)

  de <- de |>
    dplyr::mutate(
      status = dplyr::case_when(
        significant & logFC >  fc_thresh_log2 ~ "Up",
        significant & logFC < -fc_thresh_log2 ~ "Down",
        TRUE ~ "NS"
      )
    )

  label_df <- de |>
    dplyr::filter(significant) |>
    dplyr::slice_min(adj_p_value, n = 20, with_ties = FALSE)

  ggplot2::ggplot(de, ggplot2::aes(x = logFC,
                                    y = -log10(adj_p_value),
                                    colour = status)) +
    ggplot2::geom_point(alpha = 0.65, size = 1.8) +
    ggrepel::geom_label_repel(
      data    = label_df,
      mapping = ggplot2::aes(label = metabolite),
      size    = 2.5, max.overlaps = 25, box.padding = 0.3,
      show.legend = FALSE
    ) +
    ggplot2::geom_vline(xintercept = c(-fc_thresh_log2, fc_thresh_log2),
                        linetype = "dashed", colour = "grey50", linewidth = 0.5) +
    ggplot2::geom_hline(yintercept = -log10(PVAL_THRESH),
                        linetype = "dashed", colour = "grey50", linewidth = 0.5) +
    ggplot2::scale_colour_manual(
      values = c(Up = "firebrick3", Down = "steelblue4", NS = "grey75"),
      guide  = ggplot2::guide_legend(override.aes = list(size = 3))
    ) +
    ggplot2::labs(
      title   = paste("Volcano plot —", label),
      subtitle = sprintf("Model: ~ class + smoker + gender | FDR < %.2f, |log2FC| > %.3f",
                         PVAL_THRESH, fc_thresh_log2),
      x       = expression(log[2]~FC~(ADC / Healthy)),
      y       = expression(-log[10]~(FDR~adjusted~p)),
      colour  = NULL
    ) +
    ggplot2::theme_bw(base_size = 11) +
    ggplot2::theme(legend.position = "right")
}

volcano_plots <- lapply(names(de_results), function(nm) {
  make_volcano(de_results[[nm]], nm)
})
names(volcano_plots) <- names(de_results)

combined_volcano <- patchwork::wrap_plots(volcano_plots, ncol = 2)
print(combined_volcano)
save_plot("volcano_ADC1", combined_volcano, width = 14, height = 7, subdir = "DE")
```

## Heatmaps of significant metabolites

```{r heatmaps, fig.width=10, fig.height=8}
for (nm in names(de_results)) {
  de  <- de_results[[nm]]
  sig <- de |> dplyr::filter(significant) |> dplyr::pull(metabolite)

  if (length(sig) == 0) {
    message("No significant metabolites for ", nm, " — skipping heatmap.")
    next
  }

  mat_h <- t(prep_res[[nm]]$mat_de[, sig, drop = FALSE])   # features × samples
  meta_h <- prep_res[[nm]]$meta
  meta_h <- meta_h[match(colnames(mat_h), meta_h$sample_id), , drop = FALSE]

  annot_col <- data.frame(
    Class  = meta_h$class,
    Smoker = meta_h$smoker,
    Gender = meta_h$gender,
    row.names = meta_h$sample_id
  )

  cls_pal    <- c("firebrick3", "steelblue4")
  names(cls_pal) <- levels(factor(meta_h$class))
  smk_levels <- levels(factor(meta_h$smoker))
  smk_pal    <- setNames(
    RColorBrewer::brewer.pal(max(3L, length(smk_levels)), "Dark2")[seq_along(smk_levels)],
    smk_levels
  )
  gen_pal    <- c(F = "#e7298a", M = "#1f78b4")

  annot_colors <- list(Class = cls_pal, Smoker = smk_pal, Gender = gen_pal)
  annot_colors <- lapply(annot_colors, function(pal) {
    pal[names(pal) %in% levels(factor(unlist(annot_col)))]
  })

  heat_fn <- function() {
    pheatmap::pheatmap(
      mat_h,
      annotation_col    = annot_col,
      annotation_colors = annot_colors,
      scale             = "row",
      show_colnames     = FALSE,
      clustering_method = "ward.D2",
      color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
      main  = paste("Significant metabolites (row-scaled) —", nm),
      fontsize_row = max(6, 11 - length(sig) %/% 10)
    )
  }

  save_plot(paste0("heatmap_", nm), heat_fn,
            width = 10, height = max(6, length(sig) * 0.28 + 3),
            subdir = "DE")
  heat_fn()   # render inline
}
```

# Results summary

```{r results-summary}
# Overlap between Plasma and Serum significant metabolites
sig_plasma <- de_results$ADC1_Plasma |>
  dplyr::filter(significant) |> dplyr::pull(metabolite)
sig_serum  <- de_results$ADC1_Serum  |>
  dplyr::filter(significant) |> dplyr::pull(metabolite)

overlap <- intersect(sig_plasma, sig_serum)

cat(sprintf(
  "ADC1 Plasma: %d significant\nADC1 Serum:  %d significant\nOverlap:     %d (%s)\n",
  length(sig_plasma), length(sig_serum), length(overlap),
  if (length(overlap) > 0) paste(overlap, collapse = ", ") else "none"
))

# Export full combined DE table
save_table(dplyr::bind_rows(de_results), "limma_de_ADC1_combined", subdir = "DE/")
save_checkpoint(de_results, "de_results_final")
save_checkpoint(prep_res,   "prep_results_final")
```

# Session information

```{r session-info}
sessionInfo()
```
